Algebraic Groups\vspace{-1ex}

### 2019-10-14 Algebraic Geometry/Algebraic-Geometry.tex
update

>>>>

\section{October 14th, 2019}
We're going off-book a bit to talk about 
\subsection{Yoneda Lemma}
Recall that we said that $Y$ was affine if 
\[\Hom(X,Y)\cong \Hom_k(\O_Y(Y),\O_X(X)).\]
Now wer are going to take some time to put this into a broader context. Let $\calC$ be a category. We get naturally two functors 
\[h_a:\calC^{op}\to\Set\qquad h^a:\calC\to\Set\]
where 
\[h_a(b)=\Hom_\calC(b,a)\qquad h^a(b)=\Hom(a,b).\]

Now given a map $f:a\to a'$, we get natural transformations $f\circ -:h_a\to h_{a'}$ and $-\circ f:h^{a'}\to h^a$. Now notice that 
if $\ast$ is a one-point variety, then $h_X(\ast)=|X|$, the underlying point set of $X$. Notice that in a similar way 
$h_X(\A1)$ is something like the ``line space'' of $X$.

Notice that we then get a functor
\[h_{(-)}:\calC\to\mathbf{Func}(\calC^{op},\Set)\]
and we should \textbf{think} that a functor $\calC^{op}\to\Set$ is a \textbf{space over $\calC$}. That is, we can think of 
$\calC$ as the category of open sets of a topological space. We can say 
\[\Hom(U,V)=\left\{\begin{array}{lr}
	\varnothing, & U\not\subset V\\
	\{\varnothing\}, & \text{otherwise}.
\end{array}\right.\]

Notice the category $\mathbf{Open}(X)$ gives us a functor $\phi:\mathbf{Open}(X)^{op}\to\Set$ where we map 
an open $U\subseteq X$ to $\phi(U)$. For instance if $F$ is a vector bundle over $X$, we can define
\[\phi_F(U)\]
to be the sections over $U$ of the covering map. Cool.

So we're working with the slice category $\mathbf{Top}/X=\{X\to Y,\text{cts}\}$. Here morphisms are maps $Y\to X$ satisfying 
\begin{center}
	\begin{tikzcd}
		Y\ar[rr]\ar[rd] & &Z\ar[ld]\\
		& X &
	\end{tikzcd}
\end{center}

So now fix $F\to X$. Then the map $\phi_F(Y\to X)$ which is the collection of of diagrams of the form above ($Z=F$).

Then if we look at points $x\hookrightarrow X$, we get that $\phi_F(x\to X)$ is the fiber of $F$ over $x$. One ould hole that somehow 
you could recover all of $F$ from these maps, and that is precisely the content of 
\begin{lem}[Yoneda]
	The functor $h:\calC\to\mathbf{Func}(\calC^{op},\Set)$ is fully faithful.
\end{lem}
\begin{rmk}
	That is, 
	\[\Hom_\calC(a,a')\to\Hom_{\mathbf{Func}(\calC^{op},\Set)}(h_a,h_{a'})\]
	is a bijection.
\end{rmk}
\begin{prf}
	This proof is so tautological it is sometimes confusing to prove. We can look it up in any of our old notes or books but the idea is 
	to look at the image of the identity map.
\end{prf}
Next time we will see a bunch of examples and exercises.

### 2019-10-14 Algebraic Groups/Algebraic-Groups.tex
update

>>>>

\section{October 14th, 2019}
Let $G$ be a finite group. Recall the definiton of a \textbf{representation} (a linear action of $G$ on a vector space $V/k$).
This is the same data as a group homomorphism to $\GL(V)$.

\subsection{Representations of Algebraic Groups}

Now what if $G$ is an algebraic group over $k$? Now we have some exra structure of $G$ as a variety.
\begin{defn}
	A (finite dimensional) \textbf{representation} of an algebraic group $G/k$ is a homomorphism $\rho:G\to \GL(V)$ of algebraic groups.
\end{defn}
\begin{rmk}
	Notice that when $V$ is infinite-dimensional, $\GL(V)$ is no longer of finite type, so we have to let $\rho$ be a 
	morphism of group schemes.
\end{rmk}
We have the standard representation of $\GL_n$ acting on $k^n$ in the natural way. We also have the regular representation 
$G$ action on $\Gamma(G,\O_G)$. When $G=\Gm$, we get over $\bbC$ an action of $\Gm$ on $\GL(V)$ in the usual way (scaling by $\bbC^\ast$).

Observe that 
\[\rho:G\to \GL(V)=\Spec(\operatorname{Sym}^\ast(V\otimes V^\vee))_{\det}\]
corresponds to a ring morphism 
\[\Sym^\ast(V\otimes V^\vee)_{\det}\to \Gamma(G,\O_G)\eqdef \Gamma(G)\]
which corresponds to a map 
\[V\otimes V^\vee\to \Gamma(G)\]
and then tensoring with $V$, this gives us a map
\[V\xrightarrow{\sigma}\Gamma(G)\otimes V\]

So any group action gives us a \textbf{coaction} of $\Gamma(G)$ on $V$.
\begin{defn}
	A representation of $G$ is a $k$-vector space $V$ along with a coaction 
	\[\sigma:V\to \Gamma(G)\otimes V\]
	satisfying the usual dual diagrams to actions.
\end{defn}

\begin{rmk}
	As a matter of notation, recall that if $G=\Spec A$, then $A$ is a Hopf algebra. So we call $V$ an $A$-comodule.
\end{rmk}

\subsection{Reps of diagonalizable group schemes}
Let $k$ be a field (or even a ring!) and let $A$ be a finitely-generated abelian group. Define $D(A)$ to be 
\[D(A)=\oplus_{a\in A}k\cdot t^a\eqdef \Spec R\]
Then we get a multiplication 
\[R\otimes R\to R\qquad t^a\otimes t^b\mapsto t^{a+b}\]
and comultiplication
\[R\to R\otimes R\qquad t^a\to t^a\otimes t^a\]
and counit $\varepsilon$ sending $t^a\to 1$ (all $t^a$ are primitive).
\begin{prop}
	$R$ is a Hopf algebra. In particular, $D(A)\to \Spec k$ is a linear algebraic group.
\end{prop}

As an example, consider $A=\bbZ$. Then $R\cong k[t]_t$. Thus $D(A)=\Gm$.

If instead $A=\bbZ/n$, then $R\cong l[t]/(t^n-1)$, so $D(A)\cong\mu_n$.

Finally when $A\cong \bbZ^r\oplus\bbZ/n_1\oplus\cdots\oplus \bbZ/n_k$, then 
\[D(A)=\Gm^r\times\mu_{n_1}\times\cdots\times \mu_{n_k}\]

\begin{defn}
	An algebraic group over $k$ is \textbf{diagonalizable} if $G\cong D(A)$ for some $A$.
\end{defn}
Recall the definiton of irreduciblility.
\begin{thm}
	Let $A$ be af initely generated abelian group and $G=D(A)$. Then 
	\begin{itemize}
		\item Every irreducible representation of $G$ is one-dimensional and isomorphic to $I_a$, 
		corresponding to $k\to \Gamma(G)\otimes k$ where $1\mapsto t^a\otimes 1$ for some $a\in A$.
		\item Every representation decomposes as a direct sum of irreducibles.
	\end{itemize}
\end{thm}
\begin{prf}
	Let $\sigma:V\to \Gamma(G)\otimes V$ be a representation of a diagonalizable group. For $a\in A$, define 
	\[V_a\eqdef\{v\in V|\sigma(v)=t^a\otimes v\}\subseteq V\]
	Now the claim is that $V_a\cap V_b=0$ if $a\ne b$ and furthermore $\sum V_a=V$. The first isn't too hard to see.

	The second follows by considering $v\in V$ and looking at the image of it under $\sigma$. That is, 
	\[\sigma(v)=\sum_1^Nt^{\alpha_i}\otimes v_i\]
	where $\alpha_i\in A$ and $v_i\in V$. Then a very simple argument shows $v=\sum v_i$ (using linearity).
	Then it remains to show that $v_i\in V_{a_i}$, but this will make things work. (use the other axiom of a coaction).
\end{prf}
\begin{rmk}
	When $A=\bbZ$, $G=D(\bbZ)=\Gm$, which tells us that representations of $\Gm$ are in bijection with $\bbZ$-gradings 
	of $V\cong\oplus_{n\in\bbZ}V_n$!
\end{rmk}
\begin{defn}
	A linear algebraic group $G\to\Spec k$ is called \textbf{linear reductive} if every representation decomposes as a direc sum of irreducibles.
\end{defn}
\begin{prob}
	Show the above is equivalent to the statements
	\begin{itemize}
		\item for each $G$-representations $W\subseteq V$, there exists $W'\subseteq V$ subrepresentations such that $V\cong W\oplus W'$.
		\item $0\to W\to V\to W'\to 0$ is exact.
	\end{itemize} 
\end{prob}
\begin{rmk}
	Notice that this says that $D(A)$ is linear reducible. In particular, $\Gm$ and $|mu_n$ are in \textbf{any characteristic}. This runs counter to Maschke in 
	finite groups.
\end{rmk}

Consider $\bbZ/p$ in $\ch p$. We get an action $\bbZ/p$ on $k^2$ via 
\[1\cdot(x,y)=(x+y,y).\]
But notice that $k\stackrel{\hookrightarrow}{y=0}k^2$ is a subrepresentation, but has no complement! Thus this group is not linearly reductive!

As another example, consider
\[\Ga\cong\left\{\begin{pmatrix}
	1 & \alpha \\ 0 & 1
\end{pmatrix}\right\}
\subset\GL_2(k)\]
where $\Ga$ acts on $k^2$ by $\alpha(x,y)=(a+\alpha y,y)$. Then it can be easily seen not to be alinear representation.
